---
title: "Statistical findings: documentd and imputed victims and estimation of under-registration"
subtitle: "Enforced disappearance (1985-2016)"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: true
    use_bookdown: true
    highlight: kate
    embed_fonts: false
knit: (function(inputFile, encoding) {rmarkdown::render(inputFile, encoding = encoding, output_dir = "../output") })
---

```{r setup, echo = TRUE, include = FALSE}

pacman::p_load(ggplot2, dplyr, rmarkdown, LCMCR, here, arrow, dplyr, rlang, 
               purrr, glue, tidyr, stringr, gridExtra)

options(warn = -1)

```

```{r numbers, include = FALSE}

myNum <- function(n) {
  return(prettyNum(n, big.mark = " "))
}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

```{r library, include=TRUE}

library(verdata)

```

## Introduction {-}

If it's your first time working with the data, you are not familiar with the package
or you simply want to know more about the project and the objectives of these examples
and the `verdata` package, please review <!-- TODO: add link ---> before continuing.

This example illustrates the process of calculating statistics about documented,
imputed, and total victims of enforced disappearance by year. Specifically, it
replicates the upper left-hand panel of the figure found on page 12 of the [technical appendix of the project](https://www.comisiondelaverdad.co/anexo-proyecto-jep-cevhrdag) (currently only available in Spanish).

## Authenticating and importing the database (replicates) {-}

We'll begin by authenticating and importing the data about disappearances using
two functions from the `verdata` package: `confirm_files` and `read_replicates`.
Authenticating the data is relevant because the data was published with a <!-- TODO: add license -->.
This license allows the data to be distributed and modified by others. You could
have accessed these data from a number of distinct sourse, so it's important
to know whether the data have been modified.

The function `confirm_files` authenticates the files that you have downloaded.
Seeing as each violation type has 100 replicate files, this function authenticates
the data without needing to read every file into `R`. This is to save computational
resources or avoid unnecessary computation should you not want to use all 100
replicates in your analysis. This function returns a data frame with two columns:
the first column indicates the route to the fiel and the second indicates whether
the contents of the file are the same as the published version. In instances where
one or more of the files does not match the published version, the function will
also return the message _"Some replicate file contents do not match the published versions"_.

```{r confirm-replicates, eval=FALSE}

confirm <- verdata::confirm_files(here::here("verdata-parquet/desaparicion"),
                                  "desaparicion", c(1:10))

```

The function `read_replicates` allows the user to do two things: read the replicate
files into `R` in a single data frame (whether using the `.csv` or `.parquet` versions)
and verify that their contents are exactly the same as the publish data. When the
`crash` argument is set to `TRUE` (the default value), `read_replicates` will onlyy
return an object (a data frame) if the contents are equal. If the contents of any
of the files differ, `read_replicates` will return the message _"The content of the files is not identical to the ones published. This means the results of the analysis may potentially be inconsistent."_, meaning that any analyses done using these data
may result in erroneous results. <!-- TODO: add more information about behavior" -->
The data will not be read into `R`. If, for whatever reason, you want to read in
the data despite differences with the published version, you an set `crash = FALSE`.
In this case, the data will be read, but the warning will still appear. <!-- TODO: check, is this a message or a warning? -->

```{r open-replicates, echo = TRUE}

replicates_data <- verdata::read_replicates(here::here("verdata-parquet/desaparicion"),
                                            "desaparicion", c(1:10))

paged_table(replicates_data, options = list(rows.print = 10, cols.print = 5))

```

After reading in replicates 1-10, we have `r myNum(nrow(replicates_data))` records.
Additionally, we observe that the data contains information about the victim's
age, sex, and ethnicity, the year and department where the violence took place, and
the presumed perpetrator among other information. 

We will begin our analysis of enforced disappearances by transforming and filtering
this data to be suited to our question of interest.

<!-- TODO: something missing here? filter standard CEV? -->

## Documented victims {-}

After applying the necessary filters, we can now calculate statistics about the
number of documented victims disaggregated by year. These victims were documented
in the integrated database, but the records are occasionally missing information
in some fields. We will use the function `summary_observed` to calculate the number
of documented victims.

The arguments for `summary_observed` are: (1) the violation being analyzed, "desaparicion"; (2) the filtered replicate data; (3) `strata_vars`, the variables
we would like to stratify our calculations by, in this case `yy_hecho` because we
want to analyze disappearance by year; and (4) the `conflict_filter` argument which
filters to keep only people who were victims of violence in the context of the
armed conflict (variable `is_conflict == TRUE`) or not (`is_conflict == FALSE`).

`summary_observed` also has an argument called (5) `forced_dis_filter` that only
applies to analyses of enforced disappearance. This argument indicates whether
the victim was disappeared forcibly (`forced_dis == TRUE`) or not (`forced_dis == FALSE`).

There are a few additional arguments: (6) `edad_minors_filter`, which filters to
include only victims who were under the age of 18 (`edad_minors_filter = TRUE`);
(7) `include_props`, which allows for the calculation of proportions for the
variables of interest (`include_props = TRUE`) and (8) `digits`, which allows
you to control the amount of rounding for the results (by default the calculations
are rounded to two decimal places).

```{r combine-observed, echo = TRUE}

documented_table <- verdata::summary_observed("desaparicion",
                                               replicas_datos, 
                                               strata_vars = "yy_hecho",
                                               conflict_filter = TRUE,
                                               forced_dis_filter = TRUE,
                                               edad_minors_filter = FALSE,
                                               include_props = FALSE)

paged_table(documented_table, options = list(rows.print = 4, cols.print = 4))

```

```{r graph-observed, echo = TRUE}

documented_table <- documented_table %>%
    mutate(yy_hecho = as.numeric(yy_hecho)) %>% 
    arrange(desc(observed))

g <- documented_table %>%
    ggplot(aes(x = yy_hecho)) +
    geom_line(aes(y = observed, color = "Observed"),  size = 1) +
    theme_minimal() +
    theme(axis.text.x = element_text(size = 11, angle = 90),
          axis.title.y = element_text(size = 11),
          axis.ticks.x = element_line(size = 0.1)) +
    scale_x_continuous(breaks = seq(1980, 2016, 2)) +
    theme(legend.position = "bottom") +
    labs(x = "Year",
         y = "Number of victims",
         color = "") +
    scale_colour_manual(values = c("Observed" = "#434343"))

g

```

After this, we can apply the function `combine_replicates`, which will allow us
to calculate the point estimate of the mean number of victims of enforced
disappearance in the integrated database, as well as a measure of the uncertainty
of the statistical imputation process. This function uses the laws of total
expectation and variance for these calculations. 
[*Flexible Imputation of Missing Data*](https://stefvanbuuren.name/fimd/) is a
useful reference if you would like more informaion about the imputation process.

`combine_replicates` uses the following arguments: (1) the violation being analyzed,
"desaparicion"; (2) the data frame that results from applying the function
`summary_observed` to the filtered replicate data (`documented_table`); (3) the
filtered replicate data; (3) `strata_vars`, the variables
we would like to stratify our calculations by, in this case `yy_hecho` because we
want to analyze disappearance by year; and (4) the `conflict_filter` argument which
filters to keep only people who were victims of violence in the context of the
armed conflict (variable `is_conflict == TRUE`) or not (`is_conflict == FALSE`).

As was the case with `summary_observed`, `combine_replicates` also has an argument
called (5) `forced_dis_filter` that only applies to analyses of enforced 
disappearance. This argument indicates whether the victim was disappeared forcibly
(`forced_dis == TRUE`) or not (`forced_dis == FALSE`). This argument will always
be `FALSE` for other violation types.

There are a few additional arguments, again identical to those used in 
`summary_observed`: (6) `edad_minors_filter`, which filters to
include only victims who were under the age of 18 (`edad_minors_filter = TRUE`);
(7) `include_props`, which allows for the calculation of proportions for the
variables of interest (`include_props = TRUE`) and (8) `digits`, which allows
you to control the amount of rounding for the results (by default the calculations
are rounded to two decimal places).

```{r combine-replicates, echo = TRUE}

combined_table <- verdata::combine_replicates("desaparicion",
                                              tabla_documentada,
                                              replicas_datos, 
                                              strata_vars = "yy_hecho",
                                              conflict_filter = TRUE,
                                              forced_dis_filter = TRUE,
                                              edad_minors_filter = FALSE,
                                              include_props = FALSE)

paged_table(tabla_combinada, options = list(rows.print = 10, cols.print = 5))

```

```{r graph-replicates, echo = TRUE}

combined_table <- combined_table %>% 
     arrange(desc(imp_mean))

g <- combined_table %>%
    mutate(yy_hecho = as.numeric(yy_hecho)) %>% 
    ggplot(aes(x = yy_hecho)) +
    geom_line(aes(y = observed, color = "Observed"),  size = 1) +
    geom_line(aes(y = imp_mean,  color = "Imputed"), size = 1) +
    theme_minimal() +
    theme(axis.text.x = element_text(size = 11, angle = 90),
        axis.title.y = element_text(size = 11),
        axis.ticks.x = element_line(size = 0.1)) +
    scale_x_continuous(breaks = seq(1980, 2016, 2)) +
    theme(legend.position = "bottom") +
    labs(x = "Year",
         y = "Number of victims",
         color = "") +
    scale_colour_manual(values = c("Imputed" = "#1F74B1", 
                               "Observed" = "#434343"))

g

```

In this figure, the black line represents the trends in the number of documented
victims over time. Here documented victims refer to victims that we know were 
disappeared in the context of the armed conflict and who were documented as being
disappeared forcibly. That is, these victims were not missing information in the
`is_conflict` or `is_forced_dis` fields. What about records of disappearance 
victims who are missing information in one or more of these fields? The information
about these victims, what we will refer to as the imputed number of victims, is
presented alongside the number of documented victims in the blue line. This shows
the total number of victims of enforced disappearance we estimate from the victims
documented in the integrated database alone, but including victims where we were
missing key information initially, which was probabilistically "filled in" using
multiple imputation.

For example, after statistically imputing, we estimate that the 95% confidence interval for the true number of victims of enforced disappearance in the integrated database is
between `r myNum(tabla_combinada$imp_lo[1])` and `r myNum(tabla_combinada$imp_hi[1])`
for the year 2002. The point estimate of the mean number of victims during this year
is  `r myNum(tabla_combinada$imp_mean[1])`. While we often use this point estimate
to examine the tendency of the number of imputed victims over time, it's important
to remember that this statistic is incomplete without the accompanying 95% confidence
interval.

Next we will work towards estimating the total number of victims of enforced
disappearance, including those not documented by any data source.

## Stratification {-}

Stratification serves two goals: one substantive and one technical. First,
stratification allows us to specify strata—groupings of records—that are consistent
with our analytical goals. In our case, we want to analyze enforced disappearances
over time, so we will stratify our estimates by year. Second, it helps us control
capture heterogeneity, one of the key assumptions underpinning multiple systems
estimation (see the [technical appendix](https://www.comisiondelaverdad.co/anexo-proyecto-jep-cevhrdag) for more information).

To begin, we first need to filter the replicates data to include only enforced
disappearances (`is_forced_dis == 1`) and to include only disappearances that
occurred in the context of the armed conflict (`is_conflict == 1`).

```{r estratificacion, echo = TRUE}

replicate_strata <- replicates_data %>% 
  dplyr::mutate(is_conflict = as.integer(is_conflict)) %>% 
  dplyr::filter(is_conflict == 1) %>% 
  dplyr::mutate(is_forced_dis = as.integer(is_forced_dis)) %>% 
  dplyr::filter(is_forced_dis == 1)

paged_table(replicate_strata, options = list(rows.print = 10, cols.print = 5))

```

After filtering the data to include only the records consistent with our
analysis, we will stratify the data. It's important that as a data user that
you see that this process is rather artesianal, which is to say that you
can create your own code or functions to do this process. In this case, we'll
use a function called `stratify` that is not included in the `verdata` package,
but that you can copy and use in your own analyses of the data if it serves your
workflow.

```{r stratification-function, echo = TRUE}

stratify <- function(replicate_data, schema) {
    
    schema_list <- unlist(str_split(schema, pattern = ","))
    
    grouped_data <- replicate_data %>%
        group_by(!!!syms(schema_list))
    
    stratification_vars <- grouped_data %>%
        group_keys() %>%
        group_by_all() %>%
        group_split()
    
    split_data <- grouped_data %>%
        group_split(.keep = FALSE)
    
    return(list(strata_data = split_data,
                stratification_vars = stratification_vars))

}

```

The stratify function takes two arguments: (1) `replicate_data` the data to be
stratified, in our case, the filtered dataframe `replicate_strata`; and (2)
`schema`, a string with a list of variables to be used to stratify the dataframe
separate by commas.

`stratify` first groups the `replicate_data` dataframe by the variables specified
in `schema` and then splits up the grouped dataframe into a list of data frames.
Each element in the list refers to a single stratum. `stratify` returns a list.
The first element, `strata_data` is this list of data frames and the second
element, `stratification_vars` gives the specific combination of variable values 
that defines each stratum.

We can apply the function to our data in the following way:

```{r stratify-continuation, echo = TRUE}

schema <- ("replica,yy_hecho,is_forced_dis")

stratification_results <- stratify(replicate_strata, schema)

```

Let's inspect the contents of the stratification a little more closely to better
understand their contents. Here, I am going to focus on element `150` of the 
results, which corresponds to enforced disappearances in the year 2006 as
detailed by replicate 4 (`R4`).

```{r stratify-continuation2, echo = TRUE}

strata <- stratification_results[["strata_data"]]

replicate4_2006 <- strata[[150]]

paged_table(replicate4_2006, options = list(rows.print = 10, cols.print = 5))

```

Now, let's inspect the corresponding element in the `stratification_vars` part
of the result

```{r stratify-continuation3, echo = TRUE}

strata_names <- stratification_results[["stratification_vars"]]

replicate4_2006_name <- strata_names[[150]]

paged_table(replicate4_2006_name, options = list(rows.print = 10, cols.print = 5))

```

# Estimating victims by year {-}

Now that we've defined the strata we would like to estimate, we will use the
`mse` function to do the estimation using multiple systems estimation. `mse`
prepares the data to be used with the estimation functions in the `LCMCR` package,
checks if there are pre-calculated estimates available for a stratum of interest,
and fits the multiple systems estimation model if the stratum has not been
pre-calculated (or the user doesn't want to use the pre-calculated estimates).

The `mse` function takes a single stratum as an input and requires information
about the sources that an individual was documented by, the columns prefixed by
`in_`. `mse` filters these variables to include only valid lists, that is
lists that documented at least one victim in the strata of interest. That means
that `mse` will never include an `in_*` variable that takes only 0 values for
estimation, even if the original strata had one or more variables with this
structure. Additionally, `mse` only provides estimates in cases where there are
at least 3 valid sources. If a stratum is not estimable, `mse` will return an
`NA` value for the stratum.

As previously mentioned, multiple systems estimation can be time and resource
intensive. For this reason, the `mse` function allows users to make use of
pre-calculated strata used for estimates from the technical appendix. You can
download the published estimates [here](https://www.comisiondelaverdad.co/analitica-de-datos-informacion-y-recursos#c3).

Let's now take a closer look at the arguments of the `mse` function.

```{r mse-function, eval = FALSE}

mse(
  stratum_data,
  stratum_name,
  estimates_dir = NULL,
  min_n = 1,
  K = NULL,
  buffer_size = 10000,
  sampler_thinning = 1000,
  seed = 19481210,
  burnin = 10000,
  n_samples = 10000,
  posterior_thinning = 500
)

```

For this example, we'll focus on three main arguments, but you can learn more
about the rest of the arguments from the `R` help files for the package by writing
`?mse` directly into the `R` console on your computer.

- `stratum_data`: Data frame that includes data for the stratum of interest (
previously created using the `stratify` function we created earlier).

- `stratum_name`: A name to identify the stratum.

- `estimates_dir`: An optional argument, that we will use for this tutorial today.
This is the path to the directory where you have downloaded the pre-calculated
estimates. When this argument is not `NULL`, `mse` will look for the pre-calculated
estimates matching the stratum of interest in the `estimates_dir`.

```{r estimtion-files, echo = TRUE}

estimates_dir <- here::here("estimates")

```

After applying `mse` to our stratum of interest the output will be a data frame
with five columns: (1) `validated`, which indicates whether the stratum was estimable `TRUE` if yes and `FALSE` if no; (2) `N`, samples from the posterior distribution of
the probable total number of victims in the stratum, in the column (`NA`
if the stratum was not estimable); (3) `valid_sources`, the sources that were
used for the estimates; (4) `n_obs`, the number of victims documented by valid
lists in the stratum; and (5) `stratum_name`, the name of the stratum. If the stratum
is estimable, this data frame will have 1,000 rows, with 1 row per sample from
the posterior distribution. If not, the sample will only have 1 row total and 
the sample from the posterior distribution will be `NA`.

```{r estimates, eval=FALSE}

estimates <- purrr::map2_dfr(.x = stratification_results$strata_data,
                              .y = stratification_results$stratification_vars,
                              .f = verdata::mse,
                              estimates_dir = estimates_dir)

paged_table(estimates, options = list(rows.print = 10, cols.print = 8))

```

<!-- stopped here -->

```{r transformation-mse, echo = FALSE, eval = FALSE}
 
transformation <- estimates %>% 
    mutate(replicate = stratum_name$replica,
           stratum_name = paste(stratum_name$yy_hecho,sep = "-"))

# TODO: FIXME
arrow::write_parquet(transformation,
                     here::here("Resultados-CEV/Estimacion/output-estimacion/yy_hecho-desaparicion.parquet"))
 
``` 

```{r open-estimates, echo = TRUE}

estimacion_mse <- arrow::read_parquet(here::here("Resultados-CEV/Estimacion/output-estimacion/yy_hecho-desaparicion.parquet"))

paged_table(estimacion_mse, options = list(rows.print = 10, cols.print = 8))

```

```{r combinacion-transformacion, echo = TRUE}

tabla_final <- estimacion_mse %>%
    rename(replicate = replica) %>% 
    select(-validated, -valid_sources)

paged_table(tabla_final, options = list(rows.print = 10, cols.print = 5))

```